Attrition and Achievement Gaps in Online Learning

### 2015-01-24 paper_v3.Rnw
updated version

% Use this command to override the default ACM copyright statement (e.g. for preprints). 
% Consult the conference website for the camera-ready copyright statement.

%% EXAMPLE BEGIN -- HOW TO OVERRIDE THE DEFAULT COPYRIGHT STRIP -- (July 22, 2013 - Paul Baumann)
\toappear{
Permission to make digital or hard copies of all or part of this work for personal or classroom use is 	granted without fee provided that copies are not made or distributed for profit or commercial advantage and that copies bear this notice and the full citation on the first page. Copyrights for components of this work owned by others than ACM must be honored. Abstracting with credit is permitted. To copy otherwise, or republish, to post on servers or to redistribute to lists, requires prior specific permission and/or a fee. Request permissions from permissions@acm.org. \\
{\emph{L@S'15}}, March 14--15, 2015, Vancouver, Canada. \\
Copyright \copyright~2015 ACM ISBN/14/04...\$15.00. \\
DOI string from ACM form confirmation}
%% EXAMPLE END -- HOW TO OVERRIDE THE DEFAULT COPYRIGHT STRIP -- (July 22, 2013 - Paul Baumann)

% Arabic page numbers for submission. 
% Remove this line to eliminate page numbers for the camera ready copy
\pagenumbering{arabic}

@@

% Make sure hyperref comes last of your loaded packages, 
% to give it a fighting chance of not being over-written, 
% since its job is to redefine many LaTeX commands.
\usepackage[pdftex]{hyperref}
\hypersetup{
pdftitle={Attrition in Massive Open Online Courses},
pdfauthor={Rene F. Kizilcec, Sherif Halawa},
pdfkeywords={SIGCHI, proceedings, archival format},
bookmarksnumbered,
pdfstartview={FitH},
colorlinks,
citecolor=black,
filecolor=black,
linkcolor=black,
urlcolor=black,
breaklinks=true,
}

@@

@@

\begin{abstract}
Attrition in online learning is generally higher than in traditional settings, especially in large-scale online learning environments. This has prompted a shift from a monolithic notion of success toward a more nuanced treatment that accounts for the diversity in learners' motivation.
A systematic analysis of individual differences in attrition in 20 diverse massive open online courses ($N>67,000$) revealed a global achievement gap and a gender achievement gap. Women and learners in Africa, Asia, and Latin America scored lower grades and persisted less than men and learners in Europe, Oceania, and Northern America. More persistent learners were, however, only marginally more satisfied with their achievement. 
Successful online learners who left the course early reported higher levels of goal striving, feelings of social belonging, and growth mindset than unsuccessful ones ($N=756$). The shortage of time was the primary obstacle for most learners, and many were hindered by insufficient volitional control.
A better understanding of why learners disengage from MOOCs allows for earlier recognition of impending attrition, which enables targeted interventions to help learners achieve their goals.
\end{abstract}

\keywords{Massive open online course, MOOC, persistence, dropout, individual differences, volition, social belonging, mindset}

\category{H.5.m.}{Information Interfaces and Presentation (e.g. HCI)}{Miscellaneous}
\category{K.3.1.}{Computers and Education}{Computer Uses in Education}

@@

Educational environments have become increasingly diverse. Traditional schools and universities have a characteristically rigid structure, including instructor-defined---even nationally agreed---syllabi, fixed time schedules, entry requirements, and material costs to enter and exit. Novel institutional structures have been developed to overcome particular constraints. Community colleges, for instance, were created in an attempt to democratize education by offering instruction at a lower cost, with lower admission criteria, and with more flexible schedules to accommodate those who cannot afford to be full-time students \cite{goldrick2010challenges}. Distance learning programs deliver education to remote parts of the world and to people who simply could not attend in-person classes. Course materials, including assessments, were delivered through mail (correspondence education), radio, television, and eventually the Internet, thereby addressing geographical and time related constraints of traditional instruction \cite{moore1996distance}.

The latest generation of online learning environments, characterized by massive open online courses (MOOCs), has pushed the boundary on the scale of education \cite{waldrop2013campus}. By design, MOOCs provide course materials to millions of people worldwide. This scale could be achieved by pre-recording lectures, designing assessments that can be graded automatically, and by leveraging the momentum of the number of people involved (e.g., to facilitate peer learning or peer assessment). Maybe by virtue of their large scale, their prominent instructors, or their adherence to contemporary interface designs, MOOCs rapidly became an online media phenomenon. People would sign up weeks in advance of the course launch date, many of whom would never even enter the course site. And among those who enroll and enter the site, a large proportion tends to only ``sample'' some content and leave again \cite{kizilcec2013deconstructing}. Many of the prototypical behaviors observed in MOOCs \cite{kizilcec2013deconstructing,breslow2013studying} resemble those on online media platforms. This trend has also been reflected in the diversity of MOOC learners' motivations for enrolling \cite{kizilcec2015motivation}.

Shortly after the first wave of courses had finished, extensive media coverage led to MOOCs becoming associated with high attrition rates \cite{lewin2013after,parr2013mooc}. Early MOOC research cautioned against dichotomizing learners into successes and failures based on course completion \cite{kizilcec2013deconstructing,rivard2013measuring}. Instead, more nuanced categorizations based on learner behavior \cite{kizilcec2013deconstructing,clow2013moocs}, motivations \cite{kizilcec2015motivation}, or intentions \cite{wilkowski2014student} have been suggested. Ultimately, perspectives on persistence and attrition in MOOCs depend on how MOOCs have been conceptualized. For instance, MOOCs can be thougth to have bridged two worlds: one is governed by the user-centric norms of online media, where everyone is encouraged to be as active as they wish; the other adheres to the ``grammar of schooling'' \cite{tyack1994grammar}, which presupposes instructor-defined goals that students strive to achieve \cite{kizilcec2015motivation}. This view of MOOC participation clearly adds complexity to interpretations of attrition. The present work extends the empirical base of research on attrition to inform the theoretical discourse and pave the path for novel practices.

This paper presents a systematic investigation of attrition in online learning based on self-report and behavioral data collected from over 100,000 learners in 21 courses in total. Identifying characteristics associated with attrition and reasons for disengaging can facilitate earlier recognition of learners struggeling to achieve their learning goals. These learners could benefit from changes to the course or targeted interventions. We briefly review the large literature on attrition in educational environments, with a focus on key developments in understanding the causes of attrition. Building on this review, Study 1 offers insights into reasons for disengaging from MOOCs, learners' satisfaction, and individual differences in attrition and performance. For Study 2, we invited around 6,500 learners who were predicted to have disengaged from a MOOC to provide feedback. We examined differences in psychological factors between subjectively successful and unsuccessful learners, and investigated the role of volitional control for those who reported having not enough time. Learners' open responses about challenges they faced while taking the course were iteratively coded to develop an empirically-grounded list of reasons for attrition.

@@

\begin{description}
  \item[RQ1] How do prior characteristics (demographics, geographic location, intentions, and prior experience) influence learners' likelihood of disengaging from the course?
  \item[H2] Successful learners exhibit a greater sense of social belonging than unsuccessful learners.
\end{description}

@@

Distance education, in contrast to traditional in-person education, attracts a different student demographic and typically provides fewer opportunities for social integration. However, students in distance learning programs tend to lead social lives outside of school, maybe working part-time and living with their partner. Building on Tinto's model, Bean and Metzner \citeyear{bean1985conceptual} proposed a conceptual model of persistence that would be more applicable to nontraditional students. Persistence is thought to be a function of background characteristics (demographics, etc.), academic and environmental variables (study habits, financial resources, work and family obligations, etc.), and academic and psychological outcomes (GPA, satisfaction, etc.). The significant change from Tinto's orginal model was the inclusion of environmental variables to account for the added complexity of nontraditional students' lives.
% (see \cite{kember1989longitudinal}, for another adaptation of Tinto's model for distance education).

@@

Individual differences in attrition and performance are illustrated in Figure \ref{fig:s1coefs} as transformed regression coefficients with 95\% confidence intervals for each milestone. The estimates represent changes in the likelihood of achieving a milestone relative to an arbitrary baseline population. The regression intercept corresponds to the following baseline group that derives from the coding of regressors: average-aged male learners in Northern America\footnote{Learners in the United States, Canada, Bermuda, Greenland.} without a bachelor's degree, without prior experience with the topic or other MOOCs, prepared to spend up to 2 hours/week on the course, and not intending to complete it. The baseline probability for each milestone was relatively high: lectures watched (20\% above 30\%, 11\% above 50\%, 4\% above 80\%), assessments attempted (37\% above 30\%, 25\% above 50\%, 18\% above 80\%), and grades (30\% above the 60$^\text{th}$, 22\% above the 80$^\text{th}$ percentile).

@@

The large scale of MOOCs has revealed a global achievement gap and a gender achievement gap in online learning. Performance and persistence in the course was substantially lower for women and learners in Africa, Asia, and Latin America than men and learners in Europe, Oceania, and Northern America. There was a good representation from each geographic region in the dataset and the courses under consideration were on diverse topics, including some in the humanities and business with high female participation rates. The strength and significance of these relationships is remarkable, in particular with several additional demographic, geographic, and preference measures in the model. The first generation of large-scale online learning, characterized by MOOCs, is apparently not immune to inequalities that have pervaded traditional educational settings. The present work cannot move beyond presenting evidence for the existence of these gaps and discussing potential causes. Of course there exists no innate feature of women or people in certain locations that is responsible for these gaps; but rather, gender and location are associated with countless individual characteristics, some of which impact academic achievement. The achievement gaps could plausibly result from differences in Internet access, language barriers, or from feelings of psychological threat, such as fears of confirming a negative stereotype \cite{spencer1999stereotype} or not belonging in the course \cite{walton2007question}. More work is needed to identify potential causal mechanisms underlying these gaps.

@@

To gain a better understanding of the attrition patterns identified in Study 1, we designed a smaller, more focused follow-up study. This study investigates reasons for attrition based on psychological measures and learners' own descriptions of the challenges they faced. We collected feedback from learners who were likely to have stopped following the course. They were identified with a predictive model and invited to complete a survey. The survey included self-report measures of psychological constructs to test hypotheses $H1, H2,$ and $H3$ on the role of goal striving, social belonging, and growth mindset in the success of online learners. Moreover, in Study 1, most learners indicated that they disengaged due to a shortage of time. A variety of factors could influence the amount of time learners have at their disposal. Higher priority obligations, such as professional, academic, and personal committments are presumably large contributors. Another critical factor, especially in the context of MOOCs as largely unguided learning environments, is a learners' volitional control to engage in self-regulated learning \cite{corno2001volitional}. Low volitional control could reduce learners' ability to allocate time for the course in the presence of potential distractors of subjectively lower importance than taking the course (watching TV, etc.). We constructed the following hypotheses about the role of volition in time-related attrition:

@@

\begin{table*}[ht]
\caption{Descriptive and inferrential statistics for psychological measures by learner success}
\label{tab:psych}
\small
\center
\begin{tabular}{lcccc}
\toprule
 & $N$ & Goal Striving ($H1$) & Social Belonging ($H2$) & Growth Mindset ($H3$) \\
\midrule
\emph{Relative Progress} &  &  &  \\
\quad Successful & 241 & 3.46 $(SD=0.91)$ & 4.70 $(SD=0.74)$ &  4.58 $(SD=0.99)$ \\
\quad Unsuccessful & 515 & 2.96 $(SD=0.93)$ & 4.63 $(SD=0.70)$ & 4.38 $(SD=0.87)$ \\
 &  & $t_{201}=6.3, p<.001, d=0.55$ & $t_{60}=1.0, p>.25, d=.09$ & $t_{152}=2.4, p=.017, d=0.21$ \\
 \emph{Satisfaction} &  &  &  \\
\quad Successful & 325 & 3.43 $(SD=0.83)$ & 4.76 $(SD=0.70)$ & 4.47 $(SD=0.93)$ \\
\quad Unsuccessful & 431 & 2.88 $(SD=0.97)$ & 4.58 $(SD=0.71)$ & 4.42 $(SD=0.91)$ \\
&  & $t_{166}=7.2, p<.001, d=0.59$ & $t_{158}=3.1, p=.002, d=.26$ & $t_{175}=0.6, p>.25, d=.05$ \\
\bottomrule
\end{tabular}
\end{table*}

@@

Learners self-reported social belonging (17 items on their sense of social and academic fit \cite{walton2007question}, $M=4.66, SD=0.71, \alpha=0.86$); mindset (4 items on the stability of intelligence and talent\footnote{Adapted from http://mindsetonline.com/testyourmindset}; $M=4.44, SD=0.92, \alpha=0.77$), goal striving (4 items on motivation, perceived importance, committment, and confidence; $M=3.12, SD=0.95, \alpha=0.82$), and capacity for volitional control (one item about lower-priority distractions hindering the learner's course progress; $M=2.73, SD=1.29$).

@@

We further investigated research question $RQ3a$ about reasons for disengaging from the course, and tested hypotheses $H4ab$ about volitional control. In a second set of analyses, we tested our hypotheses about the role of psychological factors in the self-ascribed success of MOOC learners.

\subsubsection{Psychological Factors}

Hypotheses $H1$, $H2$, and $H3$ about goal striving, perceived social belonging, and mindset, respectively, were tested by comparing self-described successful with unsuccessful learners. Success was measured by relative progress and satisfaction with progress: 68\% of respondents reported using the course materials less than they would have wanted, and 57\% reported not being satisfied with their progress in the course (18\% were very or extremely dissatisfied). The two measures of success were correlated, $r=.35, t_{754}=10, p<.001$; satisfied respondents were, however, equally likely to be successful and unsuccessful in terms of their progress. Table \ref{tab:psych} provides summary statistics for each psychological measure by these two success metrics. Successful learners exhibited higher levels of goal striving according to both progress and satisfaction measures. Social belonging was significantly higher for learners who were successful in terms of satisfaction, but those who were successful in terms of progress. The pattern was reversed for growth mindset, such that learners who were successful in terms of progress (but not satisfaction) exhibited a stronger growth mindset. This pattern could reflect a difference in the underlying process by which these psychological factors affect learner behavior and perception. The observed differences are expected to be larger in the full population of learners, compared to the small, homogenous sample in this study, consisting of learners unlikely to persist.

Personal success in a MOOC was associated with three established psychological factors: goal striving, social belonging, and growth mindset. Prior work in more traditional achievement-oriented environments found reduced levels of these psychological factors to result in lower persistence and performance. The current observational findings highlight the critical role of these factors in an online learning environment. Learners with higher levels of goal striving, social belonging, and growth mindset reported greater personal success ($H1,H2,H3$). An online learner with, for instance, a strong sense of social belonging may derive greater satisfaction from the course and persist for longer than one with a weak sense of belonging. These psychological factors tend to vary across poeple based on prior experiences and they can change as a result of contextual cues. Pschological factors could thereby, at least partially, account for the individual differences in persistence and performance observed in Study 1.

The methodology used in this study can be extended to deliver timely and targeted interventions. The predictive model was successful in identifying a large number of learners who were facing challenges with taking the course. The majority of them were unsuccessful in their own terms. All three psychological factors under investigation were found to distinguish self-ascribed successful from unsuccessful learners. Recent work on psychological interventions found lasting positive effects of brief interventions involving reading and writing tasks (e.g., \cite{walton2007question}). Cultivating a growth mindset was found to help community college students persist in prior work, and our findings provide correlational evidence that subjectively successful learners hold a more fluid notion of intelligence than their less successful peers. Leveraging predictive models of persistence, brief interventions could be delivered online and targeted to learners who face particular challenges, such as low goal striving, a fixed mindset, or perceived threat to their social belonging.

@@

The majority of survey respondents (550 of 756; 73\%) reported being only somewhat or less satisfied with their progress in the course. While many of them (66\%) were substantially held back by other committments that took up time they had planned to spend on the course, only one in four (25\%) indicated being hindered by distractions of lower priority than the course (i.e. exhibit lower volitional control). These two time-related measures, which only differ in the assigned level of priority, were not significantly correlated ($r=.05, t_{548}=1.3, p=.21$). Reasons for learner attrition were measured using iteratively coded open responses. Table \ref{tab:s2reas} provides each coded reason and percentage of open responses that indicated it. This list of reasons maps partially onto the factors identified in Study 1 (general time issues, difficulty, format \& content, and goals \& expectations). A number of significant correlations between these challenges stood out: having not enough time was negatively associated with disliking the format or teaching style ($r=-.52$), needing additional course materials ($r=-.26$), finding the course uninteresting or not valuable ($r=-.45$), and facing technical difficulties ($r=-.29$). Hence, if a learner had not enough time, this tended to be the primary challenge.

To test hypotheses $H4ab$, we examined the 336 open responses that indicated having too little time. It revealed that no particular commitment was specified in about half of them (53\%). This distinction was a salient feature in the initial codebooks, but could not be coded with sufficient reliability across coders. Coders would have required additional training to avoid reading specific reasons into unspecific resposnes. We thus operationalized the distinction by string matching to capture professional, academic, and personal reasons for having not enough time.\footnote{Pattern: {\em work$|$job$|$school$|$university$|$college$|$family$|$kids$|$health}} Relative to those who specified a reason, learners who indicated too little time without specifying a reason also reported a significantly lower level of volitional control, i.e. their progress was more severely hindered by lower-prioirty distractions ($t_{331}=2.9, p=.003, d=.32$). However, the two groups did not significantly differ in how much they were held back by commitments that took up time they had planned to spend on the course ($t_{333}=-.75, p=.45$). These findings lend strong support to hypotheses $H4ab$: that is, while the large group of learners who reported having not enough time generally faced higher-priority obligations, a substantial proportion suffered from low volitional control---and reporting no specific reasons for having not enough time was indicative thereof.

The majority of learners indicated that they had not enough time: some struggled due to higher priority obligations that took up too much time, while others were hindered by low levels of volitional control ($H4a$). A learner needs to be highly self-regulated to be successful in an open learning environment, such as a MOOC. Very few learners explicitly indicated issues with volitional control in their open responses. They may not possess the metacognitive skills to consciously notice this issue or they preferred not to admit to this undesirable trait. Learners who reported a shortage of time without providing specific reasons were more likely to have insufficient volitional control ($H4b$). This could be a useful proxy for identifying issues with volitional control in future work. Overall, the results suggest that learners' capacity for volitional control can be a critical ingredient for learners' success and warrants further investigation \cite{corno2001volitional}.

>>>>

\toappear{\scriptsize Permission to make digital or hard copies of all or part of this work for personal or classroom use is granted without fee provided that copies are not made or distributed for profit or commercial advantage and that copies bear this notice and the full citation on the first page. Copyrights for components of this work owned by others than ACM must be honored. Abstracting with credit is permitted. To copy otherwise, or republish, to post on servers or to redistribute to lists, requires prior specific permission and/or a fee. Request permissions from permissions@acm.org. \\
{\emph{L@S 2015}}, March 14--18, 2015, Vancouver, BC, Canada. \\
Copyright is held by the owner/author(s). Publication rights licensed to ACM. \\
ACM 978-1-4503-3411-2/15/03…\$15.00 \\
http://dx.doi.org/10.1145/2724660.2724680}

@@

% Make sure hyperref comes last of your loaded packages, 
% to give it a fighting chance of not being over-written, 
% since its job is to redefine many LaTeX commands.
\usepackage[pdftex]{hyperref}
\hypersetup{
pdftitle={Attrition and Achievement Gaps in Online Learning},
pdfauthor={Rene F. Kizilcec, Sherif Halawa},
pdfkeywords={Online education; massive open online course; MOOC; persistence; dropout; individual differences; inequality; volition; social belonging; growth mindset; goal striving},
bookmarksnumbered,
pdfstartview={FitH},
colorlinks,
citecolor=black,
filecolor=black,
linkcolor=black,
urlcolor=black,
breaklinks=true,
}

@@

\clubpenalty=9996
\widowpenalty=8500

@@

\begin{abstract}
Attrition in online learning is generally higher than in traditional settings, especially in large-scale online learning environments. A systematic analysis of individual differences in attrition and grades in 20 massive open online courses ($N>67,000$) revealed a global achievement gap and a gender achievement gap. Online learners in Africa, Asia, and Latin America scored substantially lower grades and were up to half as likely to persist than those in Europe, Oceania, and Northern America. Women also exhibited lower persistence and performance then men. Yet more persistent learners were only marginally more satisfied with their achievement. The primary obstacle for most learners was finding time for the course, which was partly related to low levels of volitional control. Self-ascribed successful learners reported higher levels of goal striving, growth mindset, and feelings of social belonging than unsuccessful ones ($N=756$). 
Insights into why learners leave online courses informs models of attrition and targeted interventions to support learners achieve their goals.
\end{abstract}

\keywords{Online education; massive open online course; MOOC; persistence; dropout; individual differences; self-regulation; volition; social belonging; growth mindset; goal striving}

\category{K.3.1.}{Computers and Education}{Computer Uses in Education}
\category{H.5.m.}{Information Interfaces and Presentation (e.g. HCI)}{Miscellaneous}

@@

Educational environments have become increasingly diverse in format. Traditional schools and universities have a characteristically rigid structure, including instructor-defined syllabi, fixed time schedules, entry requirements, and material costs to enter and exit. Novel institutional structures have been developed to overcome particular constraints. Community colleges, for instance, were created in an attempt to democratize education by offering instruction at a lower cost, with lower admission criteria, and with more flexible schedules to accommodate those who cannot afford to be full-time students \cite{goldrick2010challenges}. Distance learning programs deliver education to remote parts of the world and to people who simply could not attend in-person classes. Course materials, including assessments, were delivered through mail (correspondence education), radio, television, and eventually the Internet, thereby addressing geographical and time related constraints of traditional instruction \cite{moore1996distance}.

The latest generation of online learning environments, characterized by massive open online courses (MOOCs), has pushed the boundary on the scale of education \cite{waldrop2013campus}. By design, MOOCs provide course materials to millions of people worldwide. This scale could be achieved by pre-recording lectures, designing assessments that can be graded automatically, and by leveraging the momentum of the number of people involved (e.g., to facilitate peer learning or peer assessment). Maybe by virtue of their large scale, their prominent instructors, or their adherence to contemporary interface designs, MOOCs rapidly became an online media phenomenon. People would sign up weeks in advance of the course launch date, many of whom would never even enter the course site. And among those who enroll and enter the site, a large proportion tends to only ``sample'' some content and leave again \cite{kizilcec2013deconstructing}. Many of the prototypical behaviors observed in MOOCs resemble those on online media platforms \cite{kizilcec2013deconstructing,breslow2013studying}. This trend is also reflected in the large spectrum of learners' motivations for enrolling in MOOCs \cite{kizilcec2015motivation}.

Shortly after the first wave of courses had finished, extensive media coverage led to MOOCs becoming associated with high attrition rates \cite{lewin2013after,parr2013mooc}. Early MOOC research cautioned against dichotomizing learners into successes and failures based on course completion \cite{kizilcec2013deconstructing,rivard2013measuring}. Instead, more nuanced categorizations based on learner behavior \cite{kizilcec2013deconstructing,clow2013moocs}, motivations \cite{kizilcec2015motivation}, or intentions \cite{wilkowski2014student} have been suggested. Perspectives on attrition in open online courses ultimately depend on how these courses are viewed. For instance, MOOCs can be thougth to have bridged two worlds: one is governed by the user-centric norms of online media, where everyone is encouraged to be as active as they wish; the other adheres to the ``grammar of schooling'' \cite{tyack1994grammar}, which presupposes instructor-defined goals that students strive to achieve \cite{kizilcec2015motivation}. This view of MOOC participation clearly adds complexity to interpretations of attrition. The present work extends the empirical base of research on attrition to inform the theoretical discourse and pave the path for novel practices.

This paper presents a systematic investigation of attrition in online learning based on self-report and behavioral data collected from over 100,000 learners in 21 courses in total. Identifying characteristics associated with attrition and reasons for disengaging can facilitate earlier recognition of learners struggeling to achieve their learning goals. These learners could benefit from changes to the course or targeted interventions. We briefly review the large literature on attrition in educational environments, with a focus on key developments in understanding the causes of attrition. Building on this review, Study 1 offers insights into reasons for disengaging from MOOCs, learners' satisfaction, and individual differences in attrition and performance. For Study 2, we invited around 6,053 learners who were predicted to have disengaged from a MOOC to provide feedback. We examined differences in psychological factors between subjectively successful and unsuccessful learners, and investigated the role of volitional control for those who reported having not enough time. Learners' open responses about challenges they faced while taking the course were iteratively coded to develop an empirically-grounded list of reasons for attrition.

@@

\begin{description}
  \item[RQ1] How are prior characteristics (demographics, geographic location, intentions, and prior experience) related to learners' likelihood of disengaging from the course?
  \item[H2] Successful learners exhibit a greater sense of social belonging than unsuccessful learners.
\end{description}

@@

Distance education, relative to traditional in-person education, tends to attract a more mature student demographic and provide fewer opportunities for social integration. Students in distance learning programs are more likely to lead social lives outside of school, maybe working part-time and living with their partner. Building on Tinto's model, Bean and Metzner \citeyear{bean1985conceptual} proposed a conceptual model of persistence that would be more applicable to nontraditional students. Persistence is thought to be a function of background characteristics (demographics, etc.), academic and environmental variables (study habits, financial resources, work and family obligations, etc.), and academic and psychological outcomes (GPA, satisfaction, etc.). The significant change from Tinto's orginal model was the inclusion of environmental variables to account for the added complexity of nontraditional students' lives.
% (see \cite{kember1989longitudinal}, for another adaptation of Tinto's model for distance education).

@@

Individual differences in attrition and performance are illustrated in Figure \ref{fig:s1coefs} as transformed regression coefficients with 95\% confidence intervals for each milestone. The estimates represent changes in the likelihood of achieving a milestone relative to an arbitrary baseline population. The regression intercept corresponds to the following baseline group that derives from the coding of regressors: average-aged male learners in Northern America\footnote{Learners in the United States, Canada, Bermuda, Greenland.} without a bachelor's degree, without prior experience with the topic or other MOOCs, prepared to spend up to 2 hours/week on the course, and not intending to complete it. This choice notably corresponds to the dominant power structure narrative, but is solely statistically motivated. The baseline probability for each milestone was as follows: lectures watched (20\% above 30\%, 11\% above 50\%, 4\% above 80\%), assessments attempted (37\% above 30\%, 25\% above 50\%, 18\% above 80\%), and grades (30\% above the 60$^\text{th}$, 22\% above the 80$^\text{th}$ percentile).

@@

The large scale of MOOCs has revealed a global achievement gap and a gender achievement gap in online learning. Performance and persistence in the course was substantially lower for women and learners in Africa, Asia, and Latin America than men and learners in Europe, Oceania, and Northern America. There was a good representation from each geographic region in the dataset and the courses under consideration were on diverse topics, including some in the humanities and business with high female participation rates. The strength and significance of these relationships is remarkable, in particular with several additional demographic, geographic, and preference measures in the model. The first generation of large-scale online learning, characterized by MOOCs, is apparently not immune to inequalities that have pervaded traditional educational settings. The present work presents evidence for the existence of these gaps, but cannot explain why they occur. Of course there exists no innate feature of women or people in certain locations that is responsible for these gaps. The achievement gaps could plausibly result from differences in Internet access, language barriers, or from feelings of psychological threat, such as fears of confirming a negative stereotype \cite{spencer1999stereotype} or not belonging in the course \cite{walton2007question}. A geographical categorization to compare countries with a high and low propensity of English speakers could be misleading, due to confounding with psychological barriers that are less likely to hinder learners in English-speaking countries. More work is needed to identify potential causal mechanisms underlying these gaps.

@@

To gain a better understanding of the attrition patterns identified in Study 1, we designed a smaller, more focused follow-up study. This study investigates reasons for attrition based on psychological measures and learners' own descriptions of the challenges they faced. We collected feedback from learners who were likely to have stopped following the course. They were identified with a predictive model and invited to complete a survey. The survey included self-report measures of psychological constructs to test hypotheses $H1, H2,$ and $H3$ on the role of goal striving, social belonging, and growth mindset in the success of online learners. Moreover, in Study 1, most learners indicated that they disengaged due to a shortage of time. A variety of factors could influence the amount of time learners have at their disposal. Higher priority obligations, such as professional, academic, and personal committments are presumably large contributors. Another critical factor, especially in the context of MOOCs as largely unguided learning environments, is a learner's volitional control to engage in self-regulated learning \cite{corno2001volitional}. Low volitional control could reduce learners' ability to allocate time for the course in the presence of potential distractors of subjectively lower importance than taking the course (watching TV, etc.). In contrast to attrition due to important obligations, there exist techniques to help learners become more self-regulated, which could boost their persistence. We constructed the following hypotheses about the role of volition in time-related attrition:

@@

\begin{table*}[ht]
\caption{Descriptive and inferrential statistics for psychological measures by self-ascribed success}
\label{tab:psych}
\center
\begin{tabular}{lcccc}
\toprule
 & $N$ & Goal Striving ($H1$) & Social Belonging ($H2$) & Growth Mindset ($H3$) \\
\midrule
\emph{Relative Progress} &  &  &  \\
\quad Successful & 241 & 3.46 $(SD=0.91)$ & 4.70 $(SD=0.74)$ &  4.58 $(SD=0.99)$ \\
\quad Unsuccessful & 515 & 2.96 $(SD=0.93)$ & 4.63 $(SD=0.70)$ & 4.38 $(SD=0.87)$ \\
 &  & $t_{201}=6.3, p<.001, d=0.55$ & $t_{60}=1.0, p>.25, d=.09$ & $t_{152}=2.4, p=.017, d=0.21$ \\
 \emph{Satisfaction} &  &  &  \\
\quad Successful & 325 & 3.43 $(SD=0.83)$ & 4.76 $(SD=0.70)$ & 4.47 $(SD=0.93)$ \\
\quad Unsuccessful & 431 & 2.88 $(SD=0.97)$ & 4.58 $(SD=0.71)$ & 4.42 $(SD=0.91)$ \\
&  & $t_{166}=7.2, p<.001, d=0.59$ & $t_{158}=3.1, p=.002, d=.26$ & $t_{175}=0.6, p>.25, d=.05$ \\
\bottomrule
\end{tabular}
\end{table*}

@@

Learners self-reported social belonging (17 items on their sense of social and academic fit \cite{walton2007question}, $M=4.66, SD=0.71, \alpha=0.86$); mindset (4 items on the stability of intelligence and talent\footnote{Adapted from http://mindsetonline.com/testyourmindset}; $M=4.44, SD=0.92, \alpha=0.77$), goal striving (4 items on motivation, perceived importance, committment, and confidence; $M=3.12, SD=0.95, \alpha=0.82$), and capacity for volitional control (one item on the extent to which lower-priority distractions have hindered the learner's course progress; $M=2.73, SD=1.29$).

@@

We further investigated research question $RQ3a$ about reasons for disengaging from the course, and tested hypotheses $H4ab$ about volitional control. In a second set of analyses, we tested our hypotheses about the role of psychological factors in the self-ascribed success of MOOC learners.
\\
\subsubsection{Psychological Factors}

Hypotheses $H1$, $H2$, and $H3$ about goal striving, perceived social belonging, and mindset, respectively, were tested by comparing self-ascribed successful with unsuccessful learners. Self-ascribed success was measured by relative progress and satisfaction with progress: 68\% of respondents reported using the course materials less than they would have wanted, and 57\% reported not being satisfied with their progress in the course (18\% were very or extremely dissatisfied). The two measures of success were correlated, $r=.35, t_{754}=10, p<.001$; satisfied respondents were, however, equally likely to be successful and unsuccessful in terms of their progress. Table \ref{tab:psych} provides summary statistics for each psychological measure by these two success metrics. Successful learners exhibited higher levels of goal striving according to both progress and satisfaction measures. Social belonging was significantly higher for learners who were successful in terms of satisfaction, but not those who were successful in terms of progress. The pattern was reversed for growth mindset, such that learners who were successful in terms of progress (but not satisfaction) exhibited a stronger growth mindset. This pattern could reflect a difference in the underlying process by which these psychological factors affect learner behavior and perception. The observed differences are expected to be larger in the full population of learners, compared to the homogenous sample in this study, consisting of learners who most likely left the course.

Personal success in a MOOC was associated with three established psychological factors: goal striving, social belonging, and growth mindset. Prior work in more traditional achievement-oriented environments found reduced levels of these psychological factors to result in lower persistence and performance. The current observational findings highlight the critical role of these factors in an online learning environment. Learners with higher levels of goal striving, social belonging, and growth mindset reported greater personal success ($H1,H2,H3$). An online learner with, for instance, a strong sense of social belonging may derive greater satisfaction from the course and persist for longer than one with a weak sense of belonging. These psychological factors tend to vary across people based on prior experiences and they can change as a result of contextual cues. Pschological factors could thereby, at least partially, account for the individual differences in persistence and performance observed in Study 1.

The methodology used in this study can be extended to deliver timely and targeted interventions. The predictive model was successful in identifying a large number of learners who were facing challenges with taking the course. The majority of them were unsuccessful in their own terms. All three psychological factors under investigation were found to distinguish self-ascribed successful from unsuccessful learners. Recent work on psychological interventions showed lasting positive effects of brief interventions involving reading and writing tasks (e.g., \cite{walton2007question}). Cultivating a growth mindset was found to help community college students persist in prior work, and our findings provide correlational evidence that subjectively successful learners hold a more fluid notion of intelligence than their less successful peers. Struggeling learners can be identified by leveraging predictive models of persistence. However, it requires explanatory models to decide which type of intervention should be delivered to support learners who face particular challenges, such as low goal striving, a fixed mindset, or perceived threat to their sense of belonging in the course. More research is needed to construct such explanatory models.

@@

The majority of survey respondents (550 of 756; 73\%) reported being only somewhat or less satisfied with their progress in the course. While many of them (66\%) were substantially held back by other committments that took up time they had planned to spend on the course, only one in four (25\%) indicated being hindered by distractions of lower priority than the course (a sign of lower volitional control). These two time-related measures, which only differ in the assigned level of priority, were not significantly correlated ($r=.05, t_{548}=1.3, p=.21$). Reasons for learner attrition were measured using iteratively coded open responses. Table \ref{tab:s2reas} provides each coded reason and percentage of open responses that indicated it. This list of reasons maps partially onto the factors identified in Study 1 (general time issues, difficulty, format \& content, and goals \& expectations). A number of significant correlations between these challenges stood out: having not enough time was negatively associated with disliking the format or teaching style ($r=-.52$), needing additional course materials ($r=-.26$), finding the course uninteresting or not valuable ($r=-.45$), and facing technical difficulties ($r=-.29$). Hence, if a learner had not enough time, this tended to be the primary challenge.

To test hypotheses $H4ab$, we examined the 336 open responses that indicated having too little time. It revealed that no particular commitment was specified in about half of them (53\%). This distinction was a salient feature in the initial codebooks, but could not be coded with sufficient reliability across coders. Coders would have required additional training to avoid reading specific reasons into unspecific resposnes. We thus operationalized the distinction by string matching to capture professional, academic, and personal reasons for having not enough time.\footnote{Pattern: {\em work$|$job$|$school$|$university$|$college$|$family$|$kids$|$health}} While the list of reasons is certainly not collectively exhaustive, it was based on frequently mentioned terms in the open responses. Relative to those who specified a reason, learners who indicated too little time without specifying a reason also reported significantly lower levels of volitional control, i.e. their progress was more severely hindered by lower-prioirty distractions ($t_{331}=2.9, p=.003, d=.32$). However, the two groups did not significantly differ in how much they were held back by commitments that took up time they had planned to spend on the course ($t_{333}=-.75, p=.45$). These findings lend strong support to hypotheses $H4ab$: that is, while the large group of learners who reported having not enough time generally faced higher-priority obligations, a substantial proportion appeared to be hindered by low volitional control---and reporting no specific reasons for having not enough time was indicative thereof.

The majority of learners indicated that they had not enough time: some struggled due to higher priority obligations that took up too much time, while others were hindered by low levels of volitional control ($H4a$). A learner needs to be highly self-regulated to be successful in open learning environment that provide little guidance, such as MOOCs. Very few learners explicitly indicated issues with self-regulation in their open responses. They may not possess the metacognitive skills to consciously notice the issue or they preferred not to admit to this undesirable trait. Learners who reported a shortage of time without providing specific reasons were more likely to also indicate insufficient volitional control ($H4b$). This could be a useful proxy for identifying issues of self-regulation in future work. Overall, the results suggest that additional support and guidance to build stronger self-regulatory skills could help many learners persist in the course for longer.
